Software architectures to integrate workflow engines in science gateways

### 2016-10-10 paper.tex
Updated affiliation
Started answering comments from reviewer 1

\definecolor{ao(english)}{rgb}{0.0, 0.5, 0.0}
\newcommand{\todo}[2]{\pdfmargincomment[color=red,author=#1,open=true]{#2}}
%\newcommand{\note}[2]{\pdfmargincomment[color=yellow,author=#1,open=true]{#2}}
%\newcommand{\closednote}[4]{} % closed notes are hidden. This should obviously be a parameter of \note rather than a separate command.
%\newcommand{\answerednote}[4]{\pdfmargincomment[color=green,author=#1-#3,open=true]{#1: #2 \textCR \textCR #3: #4}}
%\newcommand{\closedanswerednote}[6]{}

@@

\author[mcgill,creatis]{Tristan Glatard}
\author[mcgill]{Marc-\'Etienne Rousseau}
\author[creatis]{Sorina Camarasu-Pop}

@@

\address[mcgill]{McGill Centre for Integrative Neuroscience, Montreal Neurological Institute, McGill University, Canada.}
\address[creatis]{University of Lyon, CNRS, INSERM, CREATIS, Villeurbanne, France.}
\address[criugm]{Centre de Recherche de l'Institut de G\'eriatrie de Montr\'eal CRIUGM, Montr\'eal, QC, Canada.}
\address[isi]{University of Southern California, Information Sciences Institute, Marina del Rey, CA, USA.}
\address[amc]{Dept. Clinical Epidemiology, Biostatistics and Bioinformatics, Academic Medical Center, University of Amsterdam, NL.}
\address[stpetersburg]{St. Petersburg State University, Russia.}

@@

This paper describes, provides examples and compares such
architectures, based on system-independent representations of their
main components and interactions. It is informed by our experience in
the development and sustained operation of the CBRAIN~\cite{SHER-14},
VIP~\cite{GLAT-13} and NSG~\cite{shahand2015data,shahand:2012jgc}
science gateways for medical image analysis during the past 7 years,
as well as by lessons learned from several science gateway and
workflow projects such as
SHIWA\footnote{\url{http://www.shiwa-workflow.eu}} and
ER-flow\footnote{\url{http://www.erflow.eu}}.  In this section we
provide background information and definitions of workflow engines,
science gateways and infrastructures. In
Section~\ref{sec:architectures}, we describe six architectures that
cover the main patterns that we encountered in the various systems
that we studied during the last few years. The architectures are
described within a consistent framework that underlines the functional
interactions between their main software components. In
Section~\ref{sec:evaluation}, the different architectures are
evaluated with metrics measuring integration complexity, robustness,
extensibility, scalability and other specific features. These metrics
are specifically designed to measure the ability of the architectures
to address workflow-related issues commonly encountered in science
gateways. We finally compare the architectures and highlight
recommendations for science gateway architects and developers.

@@

  etc. Data movements can be triggered by the user in the science
  gateway (\texttt{$c_1$}), to upload input data or download processed
  data. They can also be performed by the workflow engine
  (\texttt{$c_2$}), to transfer workflow data (inputs, outputs, or
  intermediate results) across the infrastructure so that tasks can
  use them. The infrastructure might offer various data storage backends
  with heterogeneous interfaces. 
  Tools and services such as JSAGA~\cite{reynaud2010uniform} or Data Avenue~\cite{hajnal2014data}

@@

See Figure~\subref*{archi:sub-task}. The workflow engine is wrapped in
a particular task that can submit sub-tasks to the science gateway
through interaction \texttt{e}. The workflow engine keeps track of the
dependencies between the sub-tasks, but their execution is delegated
to the science gateway that executes them on the infrastructure
through interaction \texttt{b}. Although the science gateway has no
global vision of the workflow, it can keep track of the sub-tasks
submitted by a given task, for instance to cancel them when
the task is canceled. The science gateway may also implement
mechanisms to facilitate the handling of task dependencies, for
instance basic dependency lists as available in most cluster managers
(see for instance attribute \texttt{depend} of option \texttt{-W} in
Torque\footnote{\url{http://docs.adaptivecomputing.com}}).

The science gateway also transfers both user and workflow data across
the infrastructure, so that interactions \texttt{c$_1$} and
\texttt{c$_2$} are both covered by \texttt{c}.

@@

See Figure~\subref*{archi:import}. This is an abstract model
instantiated with the service invocation architecture for
consistency. Workflow engines are integrated in the science gateway
through workflow format conversion from a native format to the science
gateway format. Workflow conversion
is usually an offline process that is not involved in workflow
execution. A few systems have implemented workflow conversion. In the
SHIWA Simulation Platform, it is implemented through the IWIR
language, which provides a common language for portability across grid
workflow systems~\cite{plankensteiner-prodan-etal:2013} and allows
conversion among $n$ workflow languages using $2n$ interactions
instead of $n^2$.  Tavaxy~\cite{Abouelhoda2012} enables the import,
merging and execution of Taverna~\cite{oinn2004taverna} and
Galaxy~\cite{goecks2010galaxy} workflows; when some workflow parts
cannot be imported, workflows are run by Tavaxy as nested workflows
using their native engine. The work in~\cite{delaGarza2016} describes
workflow conversion from KNIME~\cite{Berthold2008} to WS-PGRADE/gUSE
and from Galaxy~\cite{goecks2010galaxy} to WS-PGRADE/gUSE.

@@

The architectures described in Section~\ref{sec:architectures} are
evaluated in Table~\ref{table:evaluation} using five main criteria:
integration complexity, robustness, extensibility, scalability and
other specific features. These are based on science gateway properties
introduced in Section~\ref{sec:intro_sg}. Criteria break down into
specific metrics where \emph{lower value indicates better
  performance}. Colors in Table~\ref{table:evaluation} represent
normalized scores (see caption). For each
criterion, a global score is computed by summing up the individual
scores for each metric. We ensure that the different metrics used to
calculate the global score for a criterion are comparable, so that
they can be combined meaningfully. The criteria and metrics are
explained hereafter.

@@

                                     & \cellcolor[HTML]{99A899}4

@@

                                     & \cellcolor[HTML]{99D299}\textbf{10}

@@

Science gateway \emph{integration complexity} is measured through the number of software components and their interactions. It breaks
down into the following 2 metrics:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Total components (\texttt{I$_1$}): total number of components in
  the architecture.
\item Total interactions (\texttt{I$_2$}): total number of
  interactions in the architecture.
\end{itemize}
One may wonder whether infrastructure should be counted in
\texttt{I$_1$}, since it is usually an external entity that is not developed nor maintained by the groups who
integrate workflow engines in science gateway. We support its inclusion here because integrating an
infrastructure into the system usually requires some technical
effort (e.g., account creation, software installation, APIs, etc.), therefore increasing complexity. The two metrics are summed to obtain a
global score that measures the
total number of software pieces affected by integration.

\emph{Robustness of workflow execution} measures the likelihood that the
workflow execution fails due to errors in the components or in the
interactions of the software architecture. Errors coming
from the infrastructure (e.g., unavailable data or terminated tasks)
or workflows (e.g., wrong user input or application errors) are not
covered, since they do not stem directly from the software
architecture. 
Robustness is measured here as a consequence of global
complexity, since complex architectures tend to be more prone to
failure. More precisely, robustness is determined by the number of
software components and interactions represented in
red in the architecture diagrams in
Figure~\ref{fig:architectures}, which are specific to workflow
execution. Two metrics are used:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Specific components  (\texttt{R$_1$}): number of specific components
involved in workflow executions. The Science
gateway component, for instance, is \emph{not} specific to workflow execution
since it is used to authenticate users, add new workflows, transfer
user data, etc.
\item Specific interactions (\texttt{R$_2$}): number of specific
  interactions involved in workflow executions. Data control
  (interaction \texttt{c}), for instance, is \emph{not} specific to
  workflow execution since it is also used by the science gateway to
  transfer user data. 
\end{itemize}
\texttt{R$_1$} and \texttt{R$_2$} are
summed to obtain the global score for this criterion, which measures
the total number of software pieces that are specifically involved in
the workflow execution.

\emph{Extensibility} measures the difficulty to replace or add
elements to the architecture. It is determined by the number of interactions and
components that need to be modified when a new element is
added. Modification of a component is required when its code needs to
be updated or recompiled (science gateway or workflow service), or
when a new piece of software has to be installed (infrastructure
only). Modification of an interaction is deemed necessary when the
parameters involved in this interaction are modified.  Extensibility
breaks down into 4 metrics depending on the type of element (engine, workflow or infrastructure) that has to
be added or replaced:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item New engine (\texttt{E$_1$}): number of interactions or
  components that need to be modified to integrate a new type of
  workflow engine in the architecture. Workflow engines belong to
  different types when they cannot be invoked using the same
  implementation of the interactions.
 \item Engine version upgrade (\texttt{E$_2$}): number of interactions
  or components modified to integrate a new version of a workflow
  engine in the architecture, assuming that another version of the
  same engine type is already available. Different versions of a
  workflow engine can be invoked using the same interaction
  implementation(s). When this is not the case, the different versions
  are considered as different engine types. Components are not treated
  equally regarding engine version updates. Components whose only
  function is to host the workflow engine, i.e. workflow services and
  agents, are not counted in E$_2$ because updates in such components
  are assumed to be straightforward. On the contrary, modifications of
  components that have other functions than wrapping the engine,
  i.e. science gateway and infrastructure, are counted in E$_2$
  because updates in these components require more effort, e.g. new
  release of the science gateway, gaining administrative privileges on
  the infrastructure, etc. In practice, E$_2$=0 when the workflow
  engine is wrapped in a service or in an agent, and E$_2$=1 when it is
  wrapped in another component.  
\item New workflow (\texttt{E$_3$}): number of interactions required
  to integrate a new workflow in the architecture. Adding a new
  workflow is a very common operation that does not require modifying
  software components or interactions, assuming that the engine type
  and version required to execute this workflow are already
  available. In most cases, adding a new workflow requires only
  interactions \texttt{a$_1$} and \texttt{a$_2$}, but \texttt{g}
  is required as well when workflow conversion is used. 
\item New infrastructure (\texttt{E$_4$}): number of interactions or
  components affected for the integration of a new type of infrastructure in
  the architecture. Adding a new infrastructure typically aims at providing more
  computing or storage power, enabling access to specific types of resources
  (e.g., GPUs, clouds), or  enforcing execution policies (e.g.,
  constrain data to remain in a particular network domain).
%\item New science gateway: \todo{Not sure if this makes sense}.
\end{itemize}
The four metrics are summed to obtain a global score that measures the
difficulty to extend the architecture. Note that some extensions may
affect several metrics in practice. For instance, adding a new type
of engine may help integrating new infrastructures when interactions
\texttt{b} and \texttt{c$_2$} are already present for the new engine
and the infrastructure.

\emph{Scalability} corresponds to the ability of an architecture to
cope with high workloads through the following mechanisms: starting multiple engine instances, distributing engines and task scheduling. 
This is measured by assessing the potential
scalability difficulties due to (partial) absence of a given feature in the architecture.  
 Three different features are identified:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Multiple engine instances (\texttt{S$_1$}): measures the
  potential for a single gateway instance to use more than one engine
  instance simultaneously. Workflow engines may require important
  amounts of resources when several workflows, or large workflows, are
  executed. At some point, it may be required for the science gateway
  to distribute the load among several engines. \texttt{S$_1$}=0 means
  that
  adding a new engine instance is inherently supported in the
  architecture. In this case, elastic engines can be implemented
  through some kind of auto-scaling mechanism to control the number of
  engine instances in the architecture. \texttt{S$_1$}=1 means that
  supporting multiple engine instances requires specific developments in the
  science gateway or workflow engine. \texttt{S$_1$}=2 means that new
  engine instances cannot be added.
\item Distributed engines (\texttt{S$_2$}): measures the potential of distributing
  the execution \emph{of a single workflow} among different engine
  instances. In our scope, this feature focuses on the capabilities of
  the architecture rather than these of the workflow
  engine. \texttt{S$_2$}=0 means that distributed engines are enabled by
  the architecture, and \texttt{S$_2$}=1 that they require specific
  developments in the workflow engine. 
\item Task scheduling (\texttt{S$_3$}): measures the complexity
  added by the architecture to implement task scheduling on
  the infrastructure. Task
  scheduling typically depends more on the implementation of specific
  algorithms in the science gateway, workflow engine and
  infrastructure than on the architecture used to integrate the
  workflow engine in the science gateway. Some architectures, however,
  complicate the task scheduling problem by introducing additional
  software layers or creating tasks with specific
  characteristics. \texttt{S$_3$}=0 means that the architecture does
  not add any additional complexity to the scheduling problem,
  and \texttt{S$_3$}=1 otherwise.
\end{itemize}
These three metrics are summed to obtain a global measure of the scalability
potential of the architecture.

@@

\paragraph{Extensibility} Adding a new type of workflow engine
requires implementing the corresponding workflow service,  modifying
interaction \texttt{d}, and  implementing interactions \texttt{b} and
\texttt{$c_2$} (\texttt{E$_1$}=4). New engine versions can be added by
updating the workflow service without modifying any interaction or
component (\texttt{E$_2$}=0). Updating the engine version in a
workflow service does not count in \texttt{E$_2$} since the only goal of this component
is to wrap the engine.  New workflows are added in the science gateway
or in the workflow engine through interactions \texttt{a$_1$} and
\texttt{a$_2$} (\texttt{E$_3$}=2). Adding a new type of infrastructure
requires updates in interactions \texttt{a$_1$}, \texttt{b},
\texttt{$c_1$} and \texttt{$c_2$} (\texttt{E$_4$}=4).

@@

Table~\ref{table:overall} provides an overall comparison between
architectures, based on the metrics in Table~\ref{table:evaluation}.
Note that this analysis is only meant for illustration purposes, since
it assumes that all criteria have equal weights while real systems
would favor some of them. Overall, task encapsulation performs a bit
better than the other architectures and nested workflows stand a bit
behind for the reasons mentioned previously.
\begin{table*}
\footnotesize
\centering
\begin{tabular}{rcccccc}
                                    & \textbf{Tight}
                                    & \textbf{Service}
                                    & \textbf{Task}
                                    & \textbf{Pool}
                                    & \textbf{Nested}
                                    & \textbf{Conversion} \\
  Integration (global score)    &
                                    \cellcolor[HTML]{99FF99}{0.00}
                                    & \cellcolor[HTML]{99E899}{0.22}
                                    & \cellcolor[HTML]{99FF99}{0.00}
                                    & \cellcolor[HTML]{99D299}{0.44}
                                    & \cellcolor[HTML]{999999}{1.00}
                                      & \cellcolor[HTML]{99DD99}{0.33}\\
Robustness (global score) &
                                \cellcolor[HTML]{99F099}{0.14}
                                    & \cellcolor[HTML]{99D399}{0.43}
                                    & \cellcolor[HTML]{99FF99}{0.00}
                                    & \cellcolor[HTML]{99B699}{0.71}
                                    & \cellcolor[HTML]{999999}{1.00}
                                    & \cellcolor[HTML]{99D399}{0.43}\\
  Extensibility (global score)  & \cellcolor[HTML]{99D299}{0.44}
                                     & \cellcolor[HTML]{99D299}{0.44}
                                     & \cellcolor[HTML]{99FF99}{0.00}
                                     & \cellcolor[HTML]{99E899}{0.22}
                                     & \cellcolor[HTML]{999999}{1.00}
                                     & \cellcolor[HTML]{99E899}{0.22}\\
Scalability (global score)  & \cellcolor[HTML]{999999}{1.00}
                                     & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{99FF99}{0.00}
                                     & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{99CC99}{0.50}\\
Specific features (global score) & \cellcolor[HTML]{99FF99}{0.00}
                                     & \cellcolor[HTML]{999999}{1.00}
                                     & \cellcolor[HTML]{99FF99}{0.00}
                                     & \cellcolor[HTML]{999999}{1.00}
                                     & \cellcolor[HTML]{99FF99}{0.00}
                                     & \cellcolor[HTML]{99FF99}{0.00}\\
                                    & \cellcolor[HTML]{99DA99}\textbf{1.6}
                                    & \cellcolor[HTML]{99B899}\textbf{2.6}
                                    & \cellcolor[HTML]{99FF99}\textbf{0.5}
                                    & \cellcolor[HTML]{99BF99}\textbf{2.4}
                                    & \cellcolor[HTML]{999999}\textbf{3.5}
                                    & \cellcolor[HTML]{99DE99}\textbf{1.5}\\
\end{tabular}
\caption{Overall evaluation. Brighter colors and lower scores indicate better performance. Scores
  are obtained by summing the normalized global scores ($m'$ values) of
  each criterion in Table~\ref{table:evaluation} and the colors are obtained as in Table~\ref{table:evaluation}. }
\label{table:overall}
\end{table*}

>>>>

\definecolor{ao(english)}{rgb}{0.0, 0.5, 0.0}
\newcommand{\todo}[2]{\pdfmargincomment[color=red,author=#1,open=true]{#2}}
%\newcommand{\note}[2]{\pdfmargincomment[color=yellow,author=#1,open=true]{#2}}
%\newcommand{\closednote}[4]{} % closed notes are hidden. This should obviously be a parameter of \note rather than a separate command.
%\newcommand{\answerednote}[4]{\pdfmargincomment[color=green,author=#1-#3,open=true]{#1: #2 \textCR \textCR #3: #4}}
%\newcommand{\closedanswerednote}[6]{}
\newcommand{\correction}[1]{\color{blue}#1\color{black}\xspace}

@@

\author[concordia,mcgill,creatis]{Tristan Glatard}
\author[mcgill]{Marc-\'Etienne Rousseau}
\author[creatis]{Sorina Camarasu-Pop}

@@

\address[concordia]{Department of Computer Science and Software Engineering, Concordia University, Montreal, Canada.}
\address[mcgill]{McGill Centre for Integrative Neuroscience, Montreal Neurological Institute, McGill University, Canada.}
\address[creatis]{University of Lyon, CNRS, INSERM, CREATIS, Villeurbanne, France.}
\address[criugm]{Centre de Recherche de l'Institut de G\'eriatrie de Montr\'eal CRIUGM, Montr\'eal, QC, Canada.}
\address[isi]{University of Southern California, Information Sciences Institute, Marina del Rey, CA, USA.}
\address[amc]{Dept. Clinical Epidemiology, Biostatistics and Bioinformatics, Academic Medical Center, University of Amsterdam, NL.}
\address[stpetersburg]{St. Petersburg State University, Russia.}

@@

This paper describes, provides examples and compares such
architectures, based on system-independent representations of their
main components and interactions. It is informed by our experience in
the development and sustained operation of the CBRAIN~\cite{SHER-14},
VIP~\cite{GLAT-13} and NSG~\cite{shahand2015data,shahand:2012jgc}
science gateways for medical image analysis during the past 7 years,
as well as by lessons learned from several science gateway and
workflow projects such as
SHIWA\footnote{\url{http://www.shiwa-workflow.eu}} and
ER-flow\footnote{\url{http://www.erflow.eu}}.

\correction{The analysis presented in the paper is intended for
experts in science gateway and workflow engine design. It is an
abstraction effort to identify and evaluate the fundamental
architectural patterns that are encountered while integrating workflow
engines and science gateways. In real systems, such patterns are often
intertwined due to considerations related to the historical and
technical context of software projects. The following work helps
identify the key patterns in systems and highlight potential
improvements.}

\correction{The reminder of this section provides}
 background information and definitions of workflow engines,
science gateways and infrastructures. In
Section~\ref{sec:architectures}, we describe six architectures that
cover the main patterns that we encountered in the various systems
that we studied during the last few years. The architectures are
described within a consistent framework that underlines the functional
interactions between their main software components. In
Section~\ref{sec:evaluation}, the different architectures are
evaluated with metrics measuring integration complexity, robustness,
extensibility, scalability and other specific features. These metrics
are specifically designed to measure the ability of the architectures
to address workflow-related issues commonly encountered in science
gateways. We finally compare the architectures and highlight
recommendations for science gateway architects and developers.

@@

  etc. Data movements can be triggered \correction{by} the science
  gateway (\texttt{$c_1$}), \correction{e.g.,} to upload input data or
  download processed data. They can also be performed by the workflow
  engine (\texttt{$c_2$}), to transfer workflow data (inputs, outputs,
  or intermediate results) across the infrastructure so that tasks can
  use them. The infrastructure might offer various data storage
  backends with heterogeneous interfaces.  Tools and services such as
  JSAGA~\cite{reynaud2010uniform} or Data Avenue~\cite{hajnal2014data}

@@

See Figure~\subref*{archi:sub-task}. The workflow engine is wrapped in
a particular task that can submit sub-tasks to the science gateway
through interaction \texttt{e}. The workflow engine keeps track of
the dependencies between the sub-tasks, but their execution is
delegated to the science gateway that executes them on the
infrastructure through interaction \texttt{b}. Although the science
gateway has no global vision of the workflow, it completely \correction{controls}
the sub-tasks. \correction{For instance, it may cancel} them when
the \correction{workflow} is canceled. The science gateway may also implement
mechanisms to facilitate the handling of task dependencies, for
instance basic dependency lists as available in most cluster managers
(see for instance attribute \texttt{depend} of option \texttt{-W} in
Torque\footnote{\url{http://docs.adaptivecomputing.com}}).

The science gateway also transfers both user and workflow data across
the infrastructure, through interaction \texttt{c$_1$}.

@@

See Figure~\subref*{archi:import}. This is an abstract model
instantiated with the service invocation architecture for
consistency. \correction{Workflow conversion is presented here as a pattern to
integrate workflow \emph{engines} in a science gateway, through
workflow format conversion from a native format to the science gateway
format.} Workflow conversion is usually an offline process that is not
involved in workflow execution. A few systems have implemented
workflow conversion. In the SHIWA Simulation Platform, it is
implemented through the IWIR language, which provides a common
language for portability across grid workflow
systems~\cite{plankensteiner-prodan-etal:2013} and allows conversion
among $n$ workflow languages using $2n$ interactions instead of $n^2$.
Tavaxy~\cite{Abouelhoda2012} enables the import, merging and execution
of Taverna~\cite{oinn2004taverna} and Galaxy~\cite{goecks2010galaxy}
workflows; when some workflow parts cannot be imported, workflows are
run by Tavaxy as nested workflows using their native engine. The work
in~\cite{delaGarza2016} describes workflow conversion from
KNIME~\cite{Berthold2008} to WS-PGRADE/gUSE and from
Galaxy~\cite{goecks2010galaxy} to WS-PGRADE/gUSE.

@@

The architectures described in Section~\ref{sec:architectures} are
evaluated in Table~\ref{table:evaluation} using five main criteria:
integration complexity, robustness, extensibility, scalability and
other specific features. These are based on science gateway properties
introduced in Section~\ref{sec:intro_sg}. Criteria break down into
specific metrics where \emph{lower value indicates better
  performance}. Colors in Table~\ref{table:evaluation} represent
normalized scores (see caption). For each criterion, a global score is
computed by summing up the individual scores for each metric. We
ensure that the different metrics used to calculate the global score
for a criterion are comparable, so that they can be combined
meaningfully. \correction{All the metrics in a criterion contribute
  equally to the global score. If this is not the case, the
  reader is invited to adjust the weights in the public spreadsheet available
  at \url{https://frama.link/LrPRsCQp} (column B of sheet ``Metrics'')
  to generate custom versions of Table~\ref{table:evaluation}.}

The criteria and metrics are
explained hereafter.

@@

                                     & \cellcolor[HTML]{99A899}\correction{3}

@@

                                     & \cellcolor[HTML]{99D299}\textbf{\correction{9}}

@@

Science gateway \emph{integration complexity} is measured through the number of software components and their interactions. It breaks
down into the following 2 metrics:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Total components (\texttt{I$_1$}): total number of components in
  the architecture.
\item Total interactions (\texttt{I$_2$}): total number of
  interactions in the architecture.
\end{itemize}
One may wonder whether infrastructure should be counted in
\texttt{I$_1$}, since it is usually an external entity that is not developed nor maintained by the groups who
integrate workflow engines in science gateway. We support its inclusion here because integrating an
infrastructure into the system usually requires some technical
effort (e.g., account creation, software installation, APIs, etc.), therefore increasing complexity. \correction{We assume that software components contribute equally to $I_1$ and that interactions also contribute equally to $I_2$. In a real system, this may not be the case due to a range of project-specific practical reasons that cannot be covered in this paper. The weights given to components and interactions can be dynamically changed in the public spreadsheet (sheet ``Components and interactions'') to generate new versions of Table~\ref{table:evaluation} as required.} The two metrics \correction{I$_1$ and I$_2$} are summed to obtain a
global score that measures the
total number of software pieces affected by integration.

\emph{Robustness of workflow execution} measures the likelihood that the
workflow execution fails due to errors in the components or in the
interactions of the software architecture. Errors coming
from the infrastructure (e.g., unavailable data or terminated tasks)
or workflows (e.g., wrong user input or application errors) are not
covered, since they do not stem directly from the software
architecture. 
Robustness is measured here as a consequence of global
complexity, since complex architectures tend to be more prone to
failure. More precisely, robustness is determined by the number of
software components and interactions represented in
red in the architecture diagrams in
Figure~\ref{fig:architectures}, which are specific to workflow
execution. Two metrics are used:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Specific components (\texttt{R$_1$}): number of specific
  components involved in workflow executions. \correction{For instance, the
  workflow service is a specific component because its main objective
  is to execute workflows. On the contrary, the science gateway is
  \emph{not} specific to workflow execution since it is used for a
  variety of other functions such as user authentication, data
  transfer, etc.}
\item Specific interactions (\texttt{R$_2$}): number of specific
  interactions involved in workflow executions. \correction{For
    instance, the pool-agent interaction (interaction \texttt{f}) is
    specific to workflow execution while data control (interaction
    \texttt{c$_1$}) is \emph{not} because it is used by the science gateway to transfer
    user data regardless of a particular workflow execution}.
\end{itemize}
\texttt{R$_1$} and \texttt{R$_2$} are
summed to obtain the global score for this criterion, which measures
the total number of software pieces that are specifically involved in
the workflow execution. \correction{The weights given to software components in R$_1$ and interactions in R$_2$ can be adjusted in the public spreadsheet.}

\emph{Extensibility} measures the difficulty to replace or add
elements to the architecture. It is determined by the number of interactions and
components that need to be modified when a new element is
added. Modification of a component is required when its code needs to
be updated or recompiled (science gateway or workflow service), or
when a new piece of software has to be installed (infrastructure
only). Modification of an interaction is deemed necessary when the
parameters involved in this interaction are modified.  Extensibility
breaks down into 4 metrics depending on the type of element (engine, workflow or infrastructure) that has to
be added or replaced:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item New engine (\texttt{E$_1$}): number of interactions or
  components that need to be modified to integrate a new type of
  workflow engine in the architecture. Workflow engines belong to
  different types when they cannot be invoked using the same
  implementation of the interactions.
 \item Engine version upgrade (\texttt{E$_2$}): number of interactions
  or components modified to integrate a new version of a workflow
  engine in the architecture, assuming that another version of the
  same engine type is already available. Different versions of a
  workflow engine can be invoked using the same interaction
  implementation(s). When this is not the case, the different versions
  are considered as different engine types. Components are not treated
  equally regarding engine version updates. Components whose only
  function is to host the workflow engine, i.e. workflow services and
  agents, are not counted in E$_2$ because updates in such components
  are assumed to be straightforward. On the contrary, modifications of
  components that have other functions than wrapping the engine,
  i.e. science gateway and infrastructure, are counted in E$_2$
  because updates in these components require more effort, e.g. new
  release of the science gateway, gaining administrative privileges on
  the infrastructure, etc. In practice, E$_2$=0 when the workflow
  engine is wrapped in a service or in an agent, and E$_2$=1 when it is
  wrapped in another component.  
\item New workflow (\texttt{E$_3$}): number of interactions required
  to integrate a new workflow in the architecture. Adding a new
  workflow is a very common operation that does not require modifying
  software components or interactions, assuming that the engine type
  and version required to execute this workflow are already
  available. In most cases, adding a new workflow requires only
  interactions \texttt{a$_1$} and \texttt{a$_2$}, but \texttt{g}
  is required as well when workflow conversion is used. 
\item New infrastructure (\texttt{E$_4$}): number of interactions or
  components affected for the integration of a new type of infrastructure in
  the architecture. Adding a new infrastructure typically aims at providing more
  computing or storage power, enabling access to specific types of resources
  (e.g., GPUs, clouds), or  enforcing execution policies (e.g.,
  constrain data to remain in a particular network domain).
%\item New science gateway: \todo{Not sure if this makes sense}.
\end{itemize}
The four metrics are summed to obtain a global score that measures the
difficulty to extend the architecture. \correction{The weights given to software components and interactions in E$_{1-4}$ can be adjusted in the public spreadsheet.} 
Note that some extensions may
affect several metrics in practice. For instance, adding a new type
of engine may help integrating new infrastructures when interactions
\texttt{b} and \texttt{c$_2$} are already present for the new engine
and the infrastructure.

\emph{Scalability} corresponds to the ability of an architecture to
cope with high workloads through the following mechanisms: starting multiple engine instances, distributing engines and task scheduling. 
This is measured by assessing the potential
scalability difficulties due to (partial) absence of a given feature in the architecture.  
 Three different features are identified:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Multiple engine instances (\texttt{S$_1$}): measures the
  \correction{ability} for a single gateway instance to use more than
  one engine instance simultaneously, \correction{\emph{each workflow
      being executed by a single engine instance}}.  Workflow engines
  may require important amounts of resources when several workflows,
  or large workflows, are executed. At some point, it may be required
  for the science gateway to distribute the load among several
  engines. \texttt{S$_1$}=0 means that adding a new engine instance is
  inherently supported in the architecture. In this case, elastic
  engines can be implemented through some kind of auto-scaling
  mechanism to control the number of engine instances in the
  architecture. \texttt{S$_1$}=1 means that supporting multiple engine
  instances requires specific developments in the science gateway or
  workflow engine. \texttt{S$_1$}=2 means that new engine instances
  cannot be added.
\item Distributed engines (\texttt{S$_2$}): measures the \correction{ability to distribute}
  the execution \emph{of a single workflow} among different engine
  instances. In our scope, this feature focuses on the capabilities of
  the architecture rather than these of the workflow
  engine. \texttt{S$_2$}=0 means that distributed engines are enabled by
  the architecture, and \texttt{S$_2$}=1 that they require specific
  developments in the workflow engine. 
\item Task scheduling (\texttt{S$_3$}): measures the complexity
  added by the architecture to implement task scheduling on
  the infrastructure. Task
  scheduling typically depends more on the implementation of specific
  algorithms in the science gateway, workflow engine and
  infrastructure than on the architecture used to integrate the
  workflow engine in the science gateway. Some architectures, however,
  complicate the task scheduling problem by introducing additional
  software layers or creating tasks with specific
  characteristics. \texttt{S$_3$}=0 means that the architecture does
  not add any additional complexity to the scheduling problem,
  and \texttt{S$_3$}=1 otherwise.
\end{itemize}
These three metrics are summed to obtain a global measure of the scalability
potential of the architecture.

@@

\paragraph{Extensibility} Adding a new type of workflow engine
requires implementing the corresponding workflow service,  modifying
interaction \texttt{d}, and  implementing interactions \texttt{b} and
\texttt{$c_2$}. \correction{However, when a new type of workflow engine is added, it is always possible to reuse either an existing available workflow service or interaction \texttt{d} (when a new workflow service has to be developed). Therefore, \texttt{E$_1$}=3}. New engine versions can be added by
updating the workflow service without modifying any interaction or
component (\texttt{E$_2$}=0). Updating the engine version in a
workflow service does not count in \texttt{E$_2$} since the only goal of this component
is to wrap the engine.  New workflows are added in the science gateway
or in the workflow engine through interactions \texttt{a$_1$} and
\texttt{a$_2$} (\texttt{E$_3$}=2). Adding a new type of infrastructure
requires updates in interactions \texttt{a$_1$}, \texttt{b},
\texttt{$c_1$} and \texttt{$c_2$} (\texttt{E$_4$}=4).

@@

Table~\ref{table:overall} provides an overall comparison between
architectures, based on the metrics in Table~\ref{table:evaluation}.
Note that this analysis is only meant for illustration purposes, since
it assumes that all criteria have equal weights while real systems
would favor some of them. \correction{Weights can be adjusted in the
  public spreadsheet at \url{https://frama.link/LrPRsCQp} (column B of
  sheet ``Metrics'') to produce a custom version of
  Table~\ref{table:overall} if required.}
